Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
3

\begin{figure}[!h] \noindent \small  \centering\begin{minipage}{.45\linewidth}
\begin{tabular}{lclr}
\T& ::= & \any          & \it top type\\
  & |   & \null         & \it null type\\
  & |   & \environment  & \it environment type \\
  & |   & \Language     & \it language type \\
  & |   & \symbol       & \it symbol type \\
  & |   & \S            & \it scalar type \\
  & |   & \V            & \it vector type \\
  & |   & \T~\k{|}\T    & \it union type \\
  & |   & \k{?} \T      & \it nullable type \\
  & |   & \FUN{A_1, ..., A_n}\T    & \it function type \\
  & |   & \STRUCT{F_1, ..., F_n}   & \it struct type \\
  & |   & \LIST\T                  & \it list type\\
  & |   & \CLASS{ID_1, ..., ID_n}  & \it class type\vspace{5pt}\\
\end{tabular}\end{minipage}   \hfill
\begin{minipage}{.45\linewidth}\begin{tabular}{lclr}
\A & ::= & \T          & \it arguments \\
   & |   & \VARGS      & \vspace{5pt} \\
\F & ::= & \ID:~~\T & \it struct field\vspace{5pt}\\
\V & ::= & ~\VEC\S     & \\
   & |   & \NAVEC\S    & \it na vector types \vspace{5pt}\\
\S & ::= & \int        & \\
   & |   & \chr \\
   & |   & \dbl \\
   & |   & \lgl \\
   & |   & \clx \\
   & |   & \raw \vspace{5pt}\\
\end{tabular}\end{minipage}\caption{The R type language}\label{types}
\end{figure}

@@

@@

\paragraph{Environments}

Environments in R are essentially lists with reference semantics: mutating a
value in an environment is performed in-place, and that mutation can thus be
shared with other code holding a reference to the environment.  They are
used primarily by the R runtime to store variables (their primary function
is that of an execution environment), though they also often used by
programmers either to themselves reflect on local, previous, or global
scope, or simply because of their reference semantics.  Environments are
comprised of name, value pairs, and each environment has a reference to its
parent environment.

We thought to include detailed environment types in our type language by way
of allowing environment types to define names and types of values in the
environment.  Unfortunately, this led to an undue explosion in the ``size''
of our types: often, environments had hundreds of values, particularly those
which were obtained from R (by e.g. passing a reference to a local scope).
These types were difficult to parse visually and did not feel informative.
Further, environments \AT{did not occur often}, and giving them a richer
type was not worth the increase in the visual complexity of our types.

%
%
\paragraph{Structs}

%
\paragraph{Classes}

R has more than one notion of type.  The classes of function arguments are
used by R as part of its dispatch mechanism, the most prevalent of which
(\AT{we have that number, too!}) is S3 dispatch, in which R dispatches on
the (S3) class of the first argument to a function, selecting the
appropriate method to call for the class.

Classes are very prevalent in R \AT{number}, and so we include a
\code{class<class-name>} ... \AT{todo} \typetracer was outfit to report the
class of values, but some care needed to be taken so as to not overreport
class types, as {\it all} values have a class in R.  The main classes we
were interested in were matrix, array (higher dimensional matrices), and
data.frames, in addition to any user-defined classes.  In many cases
\AT{number?}, values with a user-defined class are in reality lists with
named elements, or structs.  Including class types, and having \typetracer
report them in place of structs, significantly reduced the noise in the
analysis and allowed us to gather more focused data.

When set to ignore classes, \typetracer reports \AT{type} as the type of the
function, as compared with \AT{other type} when configured to report
classes.
%
A value can have multiple classes in R.  This is common practice in the
\code{dplyr} package, which extends the functionality of data frames: a
\code{data.frame} that passes through \code{dplyr} functions gains two
additional classes (\code{tbl} and \code{tbl_df}), bringing the total to at
least 3, in addition to whichever other classes that data frame happened to
have.  Note that when looking to perform S3 dispatch, R will try to match
{\it any} of the classes of the first argument to a dispatched function, so
each of these classes are important.

%
\paragraph{Nullable Types}

One final, common pattern that we would like to highlight is the
co-occurrence of \code{NULL} with other types of values.  In
Section~\ref{subsec:basictypes} we discussed NA, and how NAs represent
missing data for the primitive types, and more generally R provides a
\code{NULL} value that programmers use to represent missing data for other
types.  Programmers often use \code{NULL} as a default value for function
arguments, and the code snippet in Figure~\ref{fig:null} exemplifies this.

To capture this behaviour, we introduce a nullable type \code{? T} into our
type language and contract checking framework.  That way, programmers can
succinctly ascribe types to null-by-default arguments (such as in the
example in Figure~\ref{fig:null}).  This also served to reduce the number of
polymorphic arguments we detected: Without this type, \AT{X\%} of
polymorphic argument positions were of the form \code{null | T}.  Including
a nullable type served to reduce observed polymorphism by \AT{Y\%}.

@@

%
%
%
\section{Type Language Design}
\label{sec:typesystemdesign}

Our corpora analysis reported \AT{number} call traces for over \AT{number}
functions.  We distilled these call traces into function types which
captured the dynamic behaviour of the functions.  We emphasize that this was
an iterative process: initial analysis suggested a certain set of types,
which we tried and implemented, and reanalysis suggested improvements, and
so on.  In this section, we will present the final type annotation language
design, which we believe captures the essence of R.  We will end the section
with a formal type language grammar, and leading up to that we will touch on
design decisions and weigh our choices against alternatives, speaking to the
strengths and weaknesses of our chosen design.  We begin with basic,
primitive types.

%
%
%
%
\subsection{Basic Types}
\label{subsec:basictypes}

As discussed in Section~\ref{subsubsec:backgroundtypes}, the workhorse types
in R are the vectorized primitives: integers, doubles, complex numbers,
logicals (booleans), characters, and raws (bytes).  These are {\it
  vectorized} in that they are always considered to be vectors by the R
runtime: for instance, even ``scalar'' numbers are considered to be
unit-length vectors to R.  These primitive vectors are homogeneous, in that
a vector of doubles contains only doubles.

Related to these primitive types is one of R's notions of ``nullness'': for
each of the primitives, R distinguishes a special ``NA'' value which
represents missing data.  There is an NA for each of the primitive types
(i.e. there is a double NA, integer NA, etc.).  Thus, even homogeneous
vectors can represent missing data with their appropriate NA value.

As with everything in R, there is nuance in even the simplest cases.
We will now discuss our design decisions relating to the vectorized primitives.

@@

%
%
%
%
\subsection{Lists}

Separate from the vectorized primitives in R are lists: Whereas vectors are
homogeneous and restricted to primitive types, lists can contain
arbitrarily-typed data.  Further, lists and vectors have a different syntax
for accessing elements, where lists are indexed using \code{[[} and vectors
    using \code{[}.  R even builds on lists to create a more complex data
      structure known as the {\it data frame}, an essential data structure
      to R's functionality.

We are faced with several possible designs for list types.  Chief concerns
include the heterogeneity of list elements and the presence list element
names, which allow list elements to be accessed in a struct-like manner.
The next sections highlight some important design decisions for our list
types.

%
%
\subsubsection{Relationship Between Lists and Vectors}

@@

%
%
\subsubsection{Environments}

Environments in R are essentially lists with reference semantics: mutating a
value in an environment is performed in-place, and that mutation can thus be
shared with other code holding a reference to the environment.  They are
used primarily by the R runtime to store variables (their primary function
is that of an execution environment), though they also often used by
programmers either to themselves reflect on local, previous, or global
scope, or simply because of their reference semantics.  Environments are
comprised of name, value pairs, and each environment has a reference to its
parent environment.

We thought to include detailed environment types in our type language by way
of allowing environment types to define names and types of values in the
environment.  Unfortunately, this led to an undue explosion in the ``size''
of our types: often, environments had hundreds of values, particularly those
which were obtained from R (by e.g. passing a reference to a local scope).
These types were difficult to parse visually and did not feel informative.
Further, environments \AT{did not occur often}, and giving them a richer
type was not worth the increase in the visual complexity of our types.

%
%
\subsubsection{Structs}
\label{subsec:structs}

@@

%
%
%
%
\subsection{Parametric and Higher-Order Function Types}

Functions are first-class values in R.  \typetracer found \AT{number}
functions to be higher-order, having at least one function type among their
arguments.  They are markedly more common in a subset of packages, such as
\AT{such as...}.  By-and-large, these higher-order functions are similar to
well-known higher-order functions like map, filter, and fold.

The truth is that higher-order function types are best served by {\it
  parametric types} and parametric polymorphism, wherein type variables are
used within a type declaration to specify type-based relationships between
function arguments.  Unfortunately, the nature of dynamic analysis is such
that \typetracer cannot soundly detect parametric polymorphism, as without a
static analysis of the function body it is impossible to say if function
execution will halt given certain argument types, and we saw some examples
of this in
Figures~\ref{fig:scalar-vector-example}~and~\ref{fig:matrix-example}.
Without parametric types, higher-order function types become noisy and fail
to tell the complete picture.

Ultimately, we configured \typetracer to report only top-level function
types (i.e., detect if a value is a function or not).  \AT{The data suggests
  that} most of the use of higher-order functions is through map-, filter-,
and apply-like functions, whose behaviour is best captured with parametric
polymorphism.  Other common uses include higher-order functions operating on
data frame columns (in packages such as \code{dplyr}), which would
necessitate a more in-depth treatment of data frame types, which as we
discussed is a huge endeavour in and of itself.  We aim to extend our type
annotation framework with parametric, higher-order, and detailed data frame
types together in future work.

%
%
%
%
\subsection{Nullable Types}

One final, common pattern that we would like to highlight is the
co-occurrence of \code{NULL} with other types of values.  In
Section~\ref{subsec:basictypes} we discussed NA, and how NAs represent
missing data for the primitive types, and more generally R provides a
\code{NULL} value that programmers use to represent missing data for other
types.  Programmers often use \code{NULL} as a default value for function
arguments, and the code snippet in Figure~\ref{fig:null} exemplifies this.

\begin{lstlisting}
# NULL default argument example. 
\end{lstlisting}

\caption{Idiomatic NULL usage in R.}
\label{fig:null}
\end{center}
\end{figure}

To capture this behaviour, we introduce a nullable type \code{? T} into our
type language and contract checking framework.  That way, programmers can
succinctly ascribe types to null-by-default arguments (such as in the
example in Figure~\ref{fig:null}).  This also served to reduce the number of
polymorphic arguments we detected: Without this type, \AT{X\%} of
polymorphic argument positions were of the form \code{null | T}.  Including
a nullable type served to reduce observed polymorphism by \AT{Y\%}.

%
%
%
%
\subsection{Final Type Grammar}

Finally, Figure~\ref{fig:types} presents our final type language design, and
the grammar of types that \contractr is equipped to handle.  \contractr will
look for a top-level type declaration file in a package, with function types
having the form $type \; ID \; T$, where $ID$ is the function name and $T$
is a function type.  These function types have the form $\langle A_1 , ...,
A_n\rangle \Rightarrow T$, where each of the $A_i$ arguments are either some
type $T_i$, or the vararg type $...$.  Recall that functions with a vararg
parameter (\code{...}) can be called with arbitrarily many values, and extra
values are lumped together in the vararg $...$ value.  Internally, the type
$...$ is implemented as a ${\bf list} \langle any \rangle$ type.

We have discussed most of these types throughout this section, with some
notable exceptions.  ${\bf expression}$, ${\bf language}$, ${\bf symbol}$,
${\bf externalptr}$, ${\bf pairlist}$, ${\bf S4}$, and ${\bf weakref}$ types
occur \AT{rarely}, and are by-and-large uninteresting to most R programmers.
\AT{Fill in details of these basic types.}  One other type we did not
discuss is the union type $T_1 \, | \, T_2$, which is supported by
\contractr and generated by \typetracer when it cannot find some way to
consolidate argument types (for example, if an argument is observed to be
${\bf chr}$ and ${\bf int}$, \typetracer will ascribe type ${\bf chr} \, |
\, {\bf int}$ to it).  And while not generated by \typetracer (recall
discussion in Section~\ref{subsec:structs}), user-defined struct types have
the form ${\bf struct}\langle F_1, ..., F_n \rangle$, with fields $F_i$
being name $:$ type pairs.

\label{subsec:typegrammar}

\begin{figure}          % Types
    \noindent           % Collapse for readability.
    \centering
    \begin{minipage}{.45\linewidth}
    \begin{flushleft}
    % Types
      $ \begin{array}{lclr}
    D & ::= & {\bf type} \; ID \; T {\bf ;}  & \text{\it type declaration}\\
    T & ::= & {\bf null} & \text{\it unit type} \\
      & |   & {\bf any} & \text{\it top type} \\
      & |   & {\bf environment} & \text{\it environment type} \\
      & |   & {\bf expression} & \text{\it expression type} \\
      & |   & {\bf language} & \text{\it language type} \\
      & |   & {\bf symbol} & \text{\it symbol type} \\
      & |   & {\bf externalptr} & \text{\it externalptr type} \\
      & |   & {\bf pairlist} & \text{\it pairlist type} \\
      & |   & {\bf s4} & \text{\it S4 type} \\
      & |   & {\bf weakref} & \text{\it weakref type} \\
      & |   & S & \text{\it scalar type} \\
      & |   & V & \text{\it vector type} \\
      & |   & \left( T \right) & \text{\it group type} \\
      & |   & T_1 \; {\bf |} \; T_2 & \text{\it union type} \\
      & |   & ? \; T & \text{\it nullable type} \\
      & |   & \langle A_1 , ..., A_n\rangle \Rightarrow T & \text{\it function type} \\
      & |   & {\bf struct}\langle F_1, ..., F_n \rangle & \text{\it struct type} \\
      & |   & {\bf list} \langle T \rangle & \text{\it list type}\\
      & |   & {\bf class}\langle ID_1, ..., ID_n \rangle & \text{\it class type}\vspace{5pt}\\
    \end{array} $
    \end{flushleft}
    \end{minipage}
    \hfill
    \begin{minipage}{.45\linewidth}
    \begin{flushright}
    $ \begin{array}{lclr}
    % Argument Types
    A & ::= & T & \text{\it argument type} \\
      & |   & {\bf ...} \vspace{5pt} \\
    % Field Types
    F & ::= & ID : T & \text{\it struct field}\vspace{5pt}\\
    % Scalar Types
    S & ::= & B & \text{\it scalar types} \\
      & |   & N \vspace{5pt}\\
    % Vector Types
    V & ::= & S \lbrack \rbrack & \text{\it vector types} \vspace{5pt}\\
    % Base Types
    B & ::= & {\bf int} & \text{\it base types} \\
      & |   & {\bf chr} \\
      & |   & {\bf dbl} \\
      & |   & {\bf lgl} \\
      & |   & {\bf clx} \\
      & |   & {\bf raw} \vspace{5pt}\\
    % NA Types
    N & ::= & {\bf \string^int} & \text{\it na types} \\
      & |   & {\bf \string^chr} \\
      & |   & {\bf \string^dbl} \\
      & |   & {\bf \string^lgl} \\
      & |   & {\bf \string^clx} \\
\end{array} $
    \end{flushright}
    \end{minipage}
    \caption{The R type annotation language.}
    \label{fig:types}
\end{figure}

%
%
%
%
%
%
\section{Evaluation}
\label{sec:evaluation}

@@

XXX% of the parameter positions are monomorphic. 
\typetracer observes XXX calls to XXX functions with XXX arguments. This
translates to XXX parameters and XXX return types. 
A parameter or return type is  from unioning all the distinct observed types of

>>>>

\begin{figure}[!h] \noindent \small  \centering\begin{minipage}{.45\linewidth}
\begin{tabular}{lclr}
\T& ::= & \any          & \it top type\\
  & |   & \null         & \it null type\\
  & |   & \environment  & \it environment type \\
  & |   & \Language     & \it language type \\
  & |   & \symbol       & \it symbol type \\
  & |   & \S            & \it scalar type \\
  & |   & \V            & \it vector type \\
  & |   & \T~\k{|}\T    & \it union type \\
  & |   & \k{?} \T      & \it nullable type \\
  & |   & \FUN{A_1, ..., A_n}\T    & \it function type \\
  & |   & \LIST\T                  & \it list type\\
  & |   & \CLASS{ID_1, ..., ID_n}  & \it class type\vspace{5pt}\\
\end{tabular}\end{minipage}   \hfill
\begin{minipage}{.45\linewidth}\begin{tabular}{lclr}
\A & ::= & \T          & \it arguments \\
   & |   & \VARGS      & \vspace{5pt} \\
\V & ::= & ~\VEC\S     & \\
   & |   & \NAVEC\S    & \it na vector types \vspace{5pt}\\
\S & ::= & \int        & \\
   & |   & \chr \\
   & |   & \dbl \\
   & |   & \lgl \\
   & |   & \clx \\
   & |   & \raw \vspace{5pt}\\
\end{tabular}\end{minipage}\caption{The R type language}\label{types}
\end{figure}

@@

\paragraph{Nullables} 
The type system has a \null type that is inhabited by a singleton
\code{NULL} value. This value is often used as a sentinel or as a stand in
for arguments that were not provided.  To capture this behavior, we
introduce a nullable type \code{?}\T.  The difference between the \NA and
\code{NULL} values is that \NA can be stored in primitive vectors.

@@

\paragraph{Environments}
Environments are lists with reference semantics: mutating a value in an
environment is performed in-place.  They are used to store variables and to
escape from the copy-on-write semantics of other data types.  Environments
are comprised of name, value pairs, and each environment has a reference to
its parent environment.

\paragraph{Classes} 
R has more than one notion of type. Values can be attributed, one important
attribute is the class of a value.  A class is a list of names that are used
to mimic object-oriented programming. The type system includes class types
writte \CLASS{\ID_1,\dots,\ID_n} A value can have multiple classes.

\paragraph{Unions} We support untagged unions of types written
$\T_1 \k{|} \T_n$.

@@

%
%
%
\section{Left overs}

SOME OF THIS STUFF CAN BE USED IN THE EVALUATION

@@

%
%
%
%
%
\subsubsection{Relationship Between Lists and Vectors}

@@

%
%
%
\subsubsection{Structs}
\label{subsec:structs}

@@

%
%
\section{Evaluation}
\label{sec:evaluation}

@@

XXX\% of the parameter positions are monomorphic. 
\typetracer observes XXX calls to XXX functions with XXX arguments. This
translates to XXX parameters and XXX return types. 
A parameter or return type is  from unioning all the distinct observed types of
